\Large Feature and order manipulations in a free recall task affect memory for current and future lists

### 2023-01-24 paper/main.tex
entered edits from BJB

\begin{abstract} \footnotesize{We perceive, interpet, and remember ongoing
experiences through the lens of our prior experiences. Inferring that we are in
one type of situation versus another can lead us to interpret the same physical
experience differently. In turn, this can affect how we focus our attention,
form expectations about what will happen next, remember what is happening now,
draw on our prior related experiences, and so on. To study these phenomena, we
asked participants to perform simple word list learning tasks. Across different
experimental conditions, we held the set of to-be-learned words constant, but
we manipulated how irrelevant visual features changed across words and lists,
along with the orders in which the words were studied. We found that these
manipulations affected not only how the participants recalled the manipulated
lists, but also how they recalled later (random) lists. Our work shows how
structure in our ongoing experiences can exert influence over how we remember
our current experiences and unrelated subsequent experiences.

@@

Experience is subjective: different people who encounter identical physical
experiences can take away very different meanings and memories. One reason is
that our subjective experiences in the moment are shaped in part the
idiosyncratic prior experiences, memories, goals, thoughts, expectations, and
emotions that we bring with us into the present moment. These factors
collectively define a \textit{context} for our experiences~\citep{Mann20}.

@@

A key mystery is whether (and how) the sorts of situation models and schemas that people
use to organize their memories of real-world experiences might map onto the
clustering effects that reflect how people organize their memories for word
lists. On one hand, situation models and clustering effects both reflect
statistical regularities in ongoing experiences. Our memory systems exploit
these regularities when generating inferences about the unobserved past and
yet-to-be-experienced future~\citep{XuEtal23, SchaTurk15, RangRitc12,
BoweEtal79, MomeEtal17}. On the other hand, the rich structure of real-world
experiences and other naturalistic stimuli that enable people to form deep and meaningful
situation models and schemas have no obvious analog in simple word lists.  Often lists
in free recall studies are explicitly \textit{designed} to be devoid of exploitable
temporal structure, for example by sorting the words in a random order~\citep{Kaha12}.

% feature-rich free recall, basic manipulation conditions, preview of findings
We designed an experimental paradigm to explore how people organize their
memories for simple stimuli (word lists) whose temporal properties change
across different ``situations,'' analogous to how the content of real-world
experiences change across different real-world situations. We asked
participants to study and freely recall a series of word lists
(Fig.~\ref{fig:exp}). Across the different conditions in the experiment, we
varied the lists' appearances and presentation orders in different ways across
lists. The studied items (words) were designed to vary along three general
dimensions: semantic (word \textit{category}, and physical \textit{size} of the
referent), lexicographic (word \textit{length} and \textit{first letter}), and
visual (font \textit{color} and the onscreen \textit{location} of each word).
In two conditions, we manipulated whether the words' appearances were fixed or
variable within each list. In six manipulation conditions, we asked
participants to study and recall eight lists whose items were sorted by a
target feature (e.g., word category). Next, we asked them to study and recall
an additional eight lists whose items had the same features, but that were
sorted in a random temporal order. We were interested in how these 
manipulations affected participants' recall behaviors on early (manipulated) lists,
as well as how order manipulations on early lists affected recall behaviors on
later (random) lists. We used two control conditions as a baseline;
in these control conditions all of the lists were sorted randomly, but we
manipulated the presence or absence of the visual features. Finally, in an
\textit{adaptive} experimental condition we used participants' recall behaviors
on early lists to manipulate, in real-time, the presentation orders of
subsequent lists. In this adaptive condition we varied the agreement between
how participants preferred to organize their memories of the studied items
versus the orders in which the items were presented.

@@

We enrolled a total of 491 Dartmouth undergraduate students across 11
experimental conditions. The conditions included two controls (feature rich,
reduced), two visual manipulation conditions (reduced (early) and reduced
(late)), six order manipulation conditions (category, size, length, first
letter, color, and location), and a final adaptive condition. Each of these
conditions are described in the \textit{Experimental design} subsection below.

@@

We aimed to run (to completion) at least 60 participants in each of the two
primary control conditions and in the adaptive condition. In all of the other
conditions we set a target enrollment of at least 30 participants. Because our
data collection procedures entailed the coordinated efforts of 12 researchers
and multiple testing rooms and computers, it was not feasible for individual
experimenters to know how many participants had been run in each experimental
condition until the relevant databases were synchronized at the end of each
working day. We also over-enrolled participants for each condition to help
ensure that we met our minimum enrollment targets even if some participants
dropped out of the study prematurely or did not show up for their testing
session. This led us to exceed our target enrollments for several conditions.
Nevertheless, we analyze all viable data in the present paper.

@@

\paragraph{Defining feature-based distances.} Sorting words according to a
given relevant feature requires first defining a distance function for
quantifying the dissimilarity between each pair of features. This function
varied according to the type of feature under consideration. Semantic features
(category and size) are \textit{categorical}. For these features, we defined a
binary distance function: two words were considered to ``match'' (i.e., have a
distance of 0) if their labels were the same (i.e., both from the same semantic
category or both of the same size). If two words' labels were different for a
given feature, we defined the words to have a distance of 1 for that feature.
Lexicographic features (length and first letter) are \textit{discrete}. For
these features we defined a discrete distance function. Specifically, we
defined the distance between two words as either the absolute difference
between their lengths, or the absolute distance between their starting letters
in the English alphabet, respectively. For example, two words that started with
the same letter would have a ``first letter'' distance of 0, and words starting
with `J' and `A' respectively would have a first letter distance of 9. Because
words' lengths and letters' positions in the alphabet are always integers,
these discrete distances always take on integer values. Finally, the visual
features (color and location) are \textit{continuous} and
\textit{multivariate}, in that each ``feature'' takes on multiple (positive)
real values. We defined the ``color'' and ``location'' distances between two
words as the Euclidean distances between their $(r, g, b)$ color or $(x, y)$
location vectors, respectively. Therefore the color and location distance
measures always take on non-negative real values (upper-bounded at 441.67 for
color, or 27~in for location, reflecting the distances between the
corresponding maximally different vectors).

\paragraph{Constructing feature-sorted lists.} Given a list of words, a
relevant feature, and each word's value(s) for that feature, we developed a
stochastic algorithm for (noisily) sorting the words. The stochastic aspect of
our sorting procedure enabled us to obtain unique orderings for each
participant. First, we choose a word uniformly and at random from the set of
candidates. Next, we compute the distances between the chosen word's feature(s)
and the corresponding feature(s) of all yet-to-be-presented words. Third, we
convert these distances (between the previously presented word's feature
values, $a$, and the candidate word's feature values, $b$) to similarity
scores: \begin{equation} \mathrm{similarity}(a, b) = \exp\{-\tau \cdot
\mathrm{distance}(a, b)\}, \label{eqn:similarity} \end{equation} where $\tau =
1$ in our implementation. We note that increasing the value of $\tau$ would
amplify the influence of similarity on order, and decreasing the value of
$\tau$ would diminish the influence of similarity on order. Also note that this
approach requires $\tau > 0$. Finally, we computed a set of normalized
similarity values by dividing the similarities by their sum: \begin{equation}
\mathrm{similarity}_{\mathrm{normalized}}(a, b) = \frac{\mathrm{similarity}(a,
b)}{\sum_{i=1}^n \mathrm{similarity}(a, i)}, \end{equation} where in the
denominator, $i$ takes on each of the $n$ feature values of the to-be-presented
words. The resulting set of normalized similarity scores sums to 1.

@@

Beyond affecting how we process and remember \textit{ongoing} experiences, what
is happening to us now can also affect how we process and remember
\textit{future} experiences. Within the framework of our study, we wondered: if
early lists are sorted along different feature dimensions, might this affect
how people remember later (random) lists? In exploring this question, we
considered both group-level effects (i.e., effects that tended to be common
across individuals) and participant-level effects (i.e., effect that were
idiosyncratic across individuals).

@@

When we (stochastically) sorted early lists along different feature dimensions,
we found several impacts on participants' memories. Sorting early lists
semantically (by word category) enhanced participants' memories for those
lists, but the effects on performance of sorting along other feature dimensions
were inconclusive. However, each order manipulation substantially affected how
participants \textit{organized} their memories of words from the ordered lists.
When we sorted lists semantically participants displayed stronger semantic
clustering; when we sorted lists lexicographically they displayed stronger
lexicographic clustering; and when we sorted lists visually they displayed
stronger visual clustering. Clustering along the unmanipulated feature
dimensions in each of these cases was unchanged.

@@

When our expectations are violated, such as when our observations disagree with
our predictions, we may perceive the ``rules'' or ``situation'' to have
changed. \textit{Event boundaries} denote abrupts change in the state of our
experience, for example when we transition from one situation to
another~\citep{RadvZack17, ZwaaRadv98}. Crossing an event boundary can impair
our memory for pre-boundary information and enhance our memory for
post-boundary information~\citep{RadvCope06, SahaKell02, MannEtal16}. Event
boundaries are also tightly associated with the notion of \textit{situation
models} and \textit{schemas}-- mental frameworks for organizing our
understanding about the rules of how we and others are likely to behave, how
events are likely to unfold over time, how different elements are likely to
interact, and so on. For example, a situation model pertaining to a particular
restaurant might set our expectations about what we are likely to experience
when we visit that restaurant (e.g., what the building will look like, how it
will smell when we enter, how crowded the restaurant is likely to be, the
sounds we are likely to hear, etc.). Similarly, as mentioned in the
\textit{Introduction}, we might learn a schema describing how events are likely
to unfold \textit{across} any sit-down restaurant-- e.g., open the door, wait
to be seated, receive a menu, decide what to order, place the order, and so on.
Situation models and schemas can help us to generalize across our experiences,
and to generate expectations about how new experiences are likely to unfold.
When those expectations are violated, we can perceive ourselves to have crossed
into a new situation.

>>>>

\begin{abstract} \footnotesize{We perceive, interpret, and remember ongoing
experiences through the lens of our prior experiences. Inferring that we are in
one type of situation versus another can lead us to interpret the same physical
experience differently. In turn, this can affect how we focus our attention,
form expectations about what will happen next, remember what is happening now,
draw on our prior related experiences, and so on. To study these phenomena, we
asked participants to perform simple word list learning tasks. Across different
experimental conditions, we held the set of to-be-learned words constant, but
we manipulated how irrelevant visual features changed across words and lists,
along with the orders in which the words were studied. We found that these
manipulations affected not only how the participants recalled the manipulated
lists, but also how they recalled later (random) lists. Our work shows how
structure in our ongoing experiences can exert influence over how we remember
our current experiences and unrelated subsequent experiences.

@@

Experience is subjective: different people who encounter identical physical
experiences can take away very different meanings and memories. One reason is
that our subjective experiences in the moment are shaped in part by the
idiosyncratic prior experiences, memories, goals, thoughts, expectations, and
emotions that we bring with us into the present moment. These factors
collectively define a \textit{context} for our experiences~\citep{Mann20}.

@@

A key mystery is whether (and how) the sorts of situation models and schemas that people
use to organize their memories of real-world experiences might map onto the
clustering effects that reflect how people organize their memories for word
lists. On one hand, situation models and clustering effects both reflect
statistical regularities in ongoing experiences. Our memory systems exploit
these regularities when generating inferences about the unobserved past and
yet-to-be-experienced future~\citep{XuEtal23, SchaTurk15, RangRitc12,
BoweEtal79, MomeEtal17}. On the other hand, the rich structure of real-world
experiences and other naturalistic stimuli that enable people to form deep and meaningful
situation models and schemas have no obvious analog in simple word lists.  Often, lists
in free recall studies are explicitly \textit{designed} to be devoid of exploitable
temporal structure, for example, by sorting the words in a random order~\citep{Kaha12}.

% feature-rich free recall, basic manipulation conditions, preview of findings
We designed an experimental paradigm to explore how people organize their
memories for simple stimuli (word lists) whose temporal properties change
across different ``situations,'' analogous to how the content of real-world
experiences change across different real-world situations. We asked
participants to study and freely recall a series of word lists
(Fig.~\ref{fig:exp}). Across the different conditions in the experiment, we
varied the lists' appearances and presentation orders in different ways across
lists. The studied items (words) were designed to vary along three general
dimensions: semantic (word \textit{category}, and physical \textit{size} of the
referent), lexicographic (word \textit{length} and \textit{first letter}), and
visual (font \textit{color} and the onscreen \textit{location} of each word).
In two conditions, we manipulated whether the words' appearances were fixed or
variable within each list. In six manipulation conditions, we asked
participants to study and recall eight lists whose items were sorted by a
target feature (e.g., word category). Next, we asked them to study and recall
an additional eight lists whose items had the same features, but that were
sorted in a random temporal order. We were interested in how these 
manipulations affected participants' recall behaviors on early (manipulated) lists,
as well as how order manipulations on early lists affected recall behaviors on
later (random) lists. We used two control conditions as a baseline;
in these control conditions all of the lists were sorted randomly, but we
manipulated the presence or absence of the visual features. Finally, in an
\textit{adaptive} experimental condition we used participants' recall behaviors
on early lists to manipulate, in real-time, the presentation orders of
subsequent lists. In this adaptive condition, we varied the agreement between
how participants preferred to organize their memories of the studied items
versus the orders in which the items were presented.

@@

We enrolled a total of 491 Dartmouth undergraduate students across 11
experimental conditions. The conditions included two controls (feature rich,
reduced), two visual manipulation conditions (reduced (early) and reduced
(late)), six order manipulation conditions (category, size, length, first
letter, color, and location), and a final adaptive condition. Each of these
conditions is described in the \textit{Experimental design} subsection below.

@@

We aimed to run (to completion) at least 60 participants in each of the two
primary control conditions and in the adaptive condition. In all of the other
conditions, we set a target enrollment of at least 30 participants. Because our
data collection procedures entailed the coordinated efforts of 12 researchers
and multiple testing rooms and computers, it was not feasible for individual
experimenters to know how many participants had been run in each experimental
condition until the relevant databases were synchronized at the end of each
working day. We also over-enrolled participants for each condition to help
ensure that we met our minimum enrollment targets even if some participants
dropped out of the study prematurely or did not show up for their testing
session. This led us to exceed our target enrollments for several conditions.
Nevertheless, we analyze all viable data in the present paper.

@@

\paragraph{Defining feature-based distances.} Sorting words according to a
given relevant feature requires first defining a distance function for
quantifying the dissimilarity between each pair of features. This function
varied according to the type of feature under consideration. Semantic features
(category and size) are \textit{categorical}. For these features, we defined a
binary distance function: two words were considered to ``match'' (i.e., have a
distance of 0) if their labels were the same (i.e., both from the same semantic
category or both of the same size). If two words' labels were different for a
given feature, we defined the words to have a distance of 1 for that feature.
Lexicographic features (length and first letter) are \textit{discrete}. For
these features we defined a discrete distance function. Specifically, we
defined the distance between two words as either the absolute difference
between their lengths, or the absolute distance between their starting letters
in the English alphabet, respectively. For example, two words that started with
the same letter would have a ``first letter'' distance of 0, and words starting
with `J' and `A' respectively would have a first letter distance of 9. Because
words' lengths and letters' positions in the alphabet are always integers,
these discrete distances always take on integer values. Finally, the visual
features (color and location) are \textit{continuous} and
\textit{multivariate}, in that each ``feature'' takes on multiple (positive)
real values. We defined the ``color'' and ``location'' distances between two
words as the Euclidean distances between their $(r, g, b)$ color or $(x, y)$
location vectors, respectively. Therefore, the color and location distance
measures always take on non-negative real values (upper-bounded at 441.67 for
color, or 27~in for location, reflecting the distances between the
corresponding maximally different vectors).

\paragraph{Constructing feature-sorted lists.} Given a list of words, a
relevant feature, and each word's value(s) for that feature, we developed a
stochastic algorithm for (noisily) sorting the words. The stochastic aspect of
our sorting procedure enabled us to obtain unique orderings for each
participant. First, we choose a word uniformly and at random from the set of
candidates. Next, we compute the distances between the chosen word's feature(s)
and the corresponding feature(s) of all yet-to-be-presented words. Third, we
convert these distances (between the previously presented word's feature
values, $a$, and the candidate word's feature values, $b$) to similarity
scores:

\begin{equation}
 \mathrm{similarity}(a, b) = \exp\{-\tau \cdot
\mathrm{distance}(a, b)\}, \label{eqn:similarity} 
\end{equation} 
where $\tau = 1$ in our implementation. We note that increasing the value of
$\tau$ would amplify the influence of similarity on order, and decreasing the
value of $\tau$ would diminish the influence of similarity on order. Also note
that this approach requires $\tau > 0$. Finally, we computed a set of
normalized similarity values by dividing the similarities by their sum:

\begin{equation}
\mathrm{similarity}_{\mathrm{normalized}}(a, b) = \frac{\mathrm{similarity}(a,
b)}{\sum_{i=1}^n \mathrm{similarity}(a, i)}, 
\end{equation} 
where in the denominator, $i$ takes on each of the $n$ feature values of the
to-be-presented words. The resulting set of normalized similarity scores sums
to 1.

@@

Beyond affecting how we process and remember \textit{ongoing} experiences, what
is happening to us now can also affect how we process and remember
\textit{future} experiences. Within the framework of our study, we wondered: if
early lists are sorted along different feature dimensions, might this affect
how people remember later (random) lists? In exploring this question, we
considered both group-level effects (i.e., effects that tended to be common
across individuals) and participant-level effects (i.e., effects that were
idiosyncratic across individuals).

@@

When we (stochastically) sorted early lists along different feature dimensions,
we found several impacts on participants' memories. Sorting early lists
semantically (by word category) enhanced participants' memories for those
lists, but the effects on performance of sorting along other feature dimensions
were inconclusive. However, each order manipulation substantially affected how
participants \textit{organized} their memories of words from the ordered lists.
When we sorted lists semantically, participants displayed stronger semantic
clustering; when we sorted lists lexicographically, they displayed stronger
lexicographic clustering; and when we sorted lists visually, they displayed
stronger visual clustering. Clustering along the unmanipulated feature
dimensions in each of these cases was unchanged.

@@

When our expectations are violated, such as when our observations disagree with
our predictions, we may perceive the ``rules'' or ``situation'' to have
changed. \textit{Event boundaries} denote abrupt changes in the state of our
experience, for example, when we transition from one situation to
another~\citep{RadvZack17, ZwaaRadv98}. Crossing an event boundary can impair
our memory for pre-boundary information and enhance our memory for
post-boundary information~\citep{RadvCope06, SahaKell02, MannEtal16}. Event
boundaries are also tightly associated with the notion of \textit{situation
models} and \textit{schemas}-- mental frameworks for organizing our
understanding about the rules of how we and others are likely to behave, how
events are likely to unfold over time, how different elements are likely to
interact, and so on. For example, a situation model pertaining to a particular
restaurant might set our expectations about what we are likely to experience
when we visit that restaurant (e.g., what the building will look like, how it
will smell when we enter, how crowded the restaurant is likely to be, the
sounds we are likely to hear, etc.). Similarly, as mentioned in the
\textit{Introduction}, we might learn a schema describing how events are likely
to unfold \textit{across} any sit-down restaurant-- e.g., open the door, wait
to be seated, receive a menu, decide what to order, place the order, and so on.
Situation models and schemas can help us to generalize across our experiences,
and to generate expectations about how new experiences are likely to unfold.
When those expectations are violated, we can perceive ourselves to have crossed
into a new situation.
